---
title: "Winning and Attendance in the WNBA"
author:
  - name: Gabriela Eastwood
    corresponding: true
    email: geastwood@smith.edu
    affiliations:
      - ref: smith
  - name: 
      given: Benjamin S. 
      family: Baumer
    orcid: 0000-0002-3279-0516
    affiliation:
      - id: smith
        name: Smith College
        department: Statistical & Data Sciences
        address: 44 College Lane
        city: Northampton, MA
        country: United States of America
        postal-code: 01063
        url: https://www.smith.edu/
        isni: 0000000119314868
    email: bbaumer@smith.edu
    url: https://www.smith.edu/people/ben-baumer
    corresponding: false
  - name: 
      given: Andrew 
      family: Zimbalist
    affiliation:
      - ref: smith
    email: azimbali@smith.edu
    url: https://www.smith.edu/people/andrew-zimbalist
    corresponding: false


keywords:
  - WNBA
  - Game Attendance
  - Player Salary
  - Winning Percentage
abstract: |
  A central question in the intersection between sports economics and sports analytics is how to quantify the effects of what happens on the court or field on a team's revenue. In general, the literature expects that stronger teams will see higher live game attendance, but previous research has not found a statistically significant relationship between winning percentage and attendance in the WNBA [@AghaBerri2023]. Our study improves upon theirs by using a rolling win count from the past 44 games played rather than a seasonal winning-percentage to predict per game attendance. This avoids look-ahead bias, as fans cannot be expected to know a team's future performance when deciding to attend a game. Using game-level data for all WNBA regular-season games from 2018–2024, we estimated an ordinary least squares model of per-game attendance with team and season fixed effects and a dummy variable for whether Caitlin Clark played. We then translated player win shares into expected attendance impacts by mapping individual contributions to team wins onto the estimated attendance–winning relationship. We chose to focus on attendance as a proxy for revenue following the research framework from Scully (1974), due to concerns about revenue data availability. We used attendance data from the “Across the Timeline: WNBA League Attendance Dashboard”, which records game-level attendance data for all WNBA games. We find a significant relationship between rolling win count and game attendance, with game attendance expected to increase by 112 people for each additional win in the 44 game window. This means that a player like Allisha Gray, for example, who we estimate contributed 10 wins to the Atlanta Dream in 2024, boosted attendance by approximately 1,100 attendees per game. These results highlight the importance of short-run performance in shaping fan demand and provide a framework for linking individual player performance to team-level revenue outcomes in women’s professional sports.
key-points:
  - winning percentage and game attendance
date: last-modified
bibliography: references.bib
number-sections: true
---
```{r}
#| message: false
library(tidyverse)
library(knitr)
library(gt)
```

```{r}
wnba_gl <- read_rds(here::here("data/wnba_gl.rds"))
```

## Introduction

Research on pay and performance in sports economics often centers on how teams value player contributions and how those valuations translate into compensation. A central challenge in the field is determining the appropriate metrics for quantifying an individual player’s impact on team success and linking that impact to team revenues and attendance. Different modeling choices can lead to significantly different conclusions about whether players are compensated fairly for their contributions to the game. As a relatively young league with rapid growth in viewership [@wnba2024recordseason] but limited publicly available financial data, accurately valuing player contributions in the WNBA is difficult. We have found attendance data to be the most reliably available indicator of revenue, making the link between on-court performance and fan demand especially important for understanding the valuation of players.

The underlying assumption of marginal revenue models is simple: fans are more interested in watching good teams. In their 2014 paper, Coates et al. use a consumer-choice model to explain what impacts a fan’s choice to attend a live MLB game. While they look specifically at baseball, it is reasonable to assume that similar considerations apply to other sports as well. Coates et al. explain the traditional model for explaining fan behavior: the uncertainty of outcome hypothesis. This is idea that fans are more likely to attend games when the outcome is uncertain, making the game more exciting. But more recent behavioral economics suggests that people evaluate gains and losses relative to a reference point and tend to dislike losses more than they like equivalent gains [@CoatesHumphreysZhou2014]. They argue that applying this idea of loss aversion may better explain why fans choose to go games than the uncertainty of outcome hypothesis. They use MLB live-game attendance data to find attendance patterns consistent with the hypothesis that fans are less likely to attend games in which they expect the home-team to lose [@CoatesHumphreysZhou2014].
  
Building on the idea that wins translate directly into fan interest and team revenue, Scully’s 1974 study on pay and performance in Major League Baseball is one of the first empirical studies on the labor market in a professional sports environment [@HolmesSimmonsBerri2018]. Scully investigates whether MLB players were undervalued relative to their estimated marginal revenue products [@Scully1974]. He estimates players’ marginal revenue products by regressing a team’s revenue on their winning percentage and computing individual players’ contributions to their teams winning percentage. This allows him to evaluate how much each player contributed to the team’s total revenue in a season, estimating their marginal revenue product and comparing it to their salary. Scully’s study came two years after the first players strike in baseball history, and aimed to evaluate whether players’ claims that they were being systematically undervalued were accurate. Scully finds that the average MLB player at the time was being paid 20% of his marginal revenue contribution, and that exploitation was higher for early career players [@Scully1974]. He also finds that certain exceptional players were being overvalued compared to their estimated marginal revenue contributions, and hypothesizes this reveals that players can generate revenue for their teams outside of their effect on game outcomes (i.e. the Caitlyn Clark effect).
  
Subsequent scholars have raised important critiques of Scully’s method. Krautmann criticizes Scully’s method by showing that the results are highly sensitive to how a player’s contribution to team performance is measured [@Krautmann1999]. Because there are many possible ways to quantify an individual’s on-field contribution, his critique highlights the need—especially for our research—to use a rigorous and defensible metric for estimating the number of wins each player contributes. In his 1991 paper on salaries and performance in MLB, Zimbalist notes that Scully’s model doesn’t account for how poor play can contribute to a team’s losses and thus may inflate the marginal revenue product of each player [@Zimbalist1992].
	
  Finally, situating these revenue models within broader league structure, sports economists have written extensively about the relationship between winningness and team revenue, particularly about how competitive balance benefits teams across the league in generating revenue. While teams and fans want to win as many games as possible, significant imbalance in winning percentages across a professional sports league will tend to decrease viewership as spectators will only watch a game if they have some uncertainty about the outcome [@PerlineNobleStoldt2018]. The various professional sports leagues in the United States have all instituted financial restrictions on teams in their leagues in the interest of competitive balance, such as across-league revenue sharing, salary caps, and reverse-of-finish drafts. The fundamental premise behind all of these attempts to protect competitive balance is that by normalizing the amounts that teams are spending on their rosters you can weaken the ability of richer teams to ‘buy wins’ and level the playing field enough to keep things interesting. This ought to increase the revenue of the whole league by encouraging attendance and engagement. In their study of competitive balance in NCAA basketball, Perline et al. found that the women’s game was significantly less balanced than the men’s game, looking at the standard deviation and the range of winning percentages for teams in the women’s division I conference [@PerlineNobleStoldt2018]. They hypothesize that the difference in the competitive balance of the men’s and women’s games could be one cause of lower revenue generation and attendance in the women’s game. The importance of class balance to the health of a league is interesting, but ultimately outside the scope of our research which seeks to quantify how players’ performance contributes to their team’s revenue in the WNBA.

In their 2023 paper comparing the in person attendance demand of the WNBA and the NBA, Agha and Berri attempt to establish whether different forces shape demand for attendance in the WNBA and the NBA. They model attendance in both leagues as a function of the population of the team’s home location, the winning percentage of the team, the per capita income of the area, championships won and player quality. Interestingly, they don’t generate statistically significant results for the effect of winning percentage on game attendance in the WNBA, while it is significant in the NBA [@AghaBerri2023]. They hypothesize that this may be due to decreased media coverage of the WNBA, and the fact that the WNBA is a much younger league, but they don’t explain why these factors would decrease the effect of winning percentage on game attendance. Although Agha and Berri did not find statistically significant effects, our method differs from theirs in two important ways. Firstly, we are regressing onto raw per game attendance counts rather than attendance as a percentage of arena capacity. Percentage of arena capacity does not directly capture attendance or game demand as arena sizes vary across teams, so raw attendance counts more effectively reflect fan demand. Secondly, we introduce a rolling win count over a 44 game window rather than using seasonal winning percentage to predict per game attendance. This avoids look-ahead bias, as fan behavior can not be reasonably expected to reflect future unseen team success. Replicating their analysis with a simpler model focused solely on win count and fixed team and seasonal effects may yield clearer evidence on how team performance influences attendance. We aim to build upon their work by combining their model with Scully’s and first regressing game attendance onto win count, and then using player statistics to estimate individual player contributions to winning percentage. This would allow us to evaluate how a player’s on-court performance contributes to game attendance, and therefore revenue. In Scully’s original 1974 model, attendance multiplied by ticket price serves as an estimate for team revenue. While our analysis cannot capture broadcast or merchandise income, game attendance remains a meaningful indicator of team visibility and revenue generation, and it is by far the most complete and publicly available dataset. Focusing on attendance as the response variable is therefore both practical and justifiable.

## Data {#sec-data}

Our analysis is based on game-level data that spans WNBA history, beginning in 1997 and continuing through the end of the 2025 season. 
Our final data set contains observations from `r format(nrow(wnba_gl), big.mark = ",")` games.

First, we collected game-level attendance data from [@AcrossTheTimeline_WNBA_Attendance]. 
We supplemented these data with arena capacities that we collected manually[^arenas]. 
Attendance data was unavailable for only `r nrow(filter(wnba_gl, is.na(attendance)))` games, while (due to the COVID-19 pandemic) `r nrow(filter(wnba_gl, attendance == 0))` games (accurately) recorded attendance of 0. 

[^arenas]: We were only able to collect an estimate for the *current* arena capacity, which is not necessarily the same as the arena capacity on the day of the game. 

Second, we collected game-level performance data from the **wehoop** R package [@hutchinson_gilani_2021_wehoop], which uses the official WNBA API. 
These data include the final score of the game. 
We also used the **wehoop** package to identify whether Caitlin Clark played more than 0 minutes in the game. 

@tbl-attendance shows a sample of the data. 
While many other variables are omitted from the table, we note that the attendance, the teams playing in the game, the outcome, the arena, and whether Caitlin Clark played in the game, are recorded. 

```{r}
#| label: tbl-attendance
#| fig-cap: "Sample of Game-level WNBA data. The data indicates (among other things) which teams were playing, who won, the arena, the attendance, and whether Caitlin Clark played in the game."

wnba_gl |>
  select(
    game_date,
    home = team_abbreviation,
    away = opponent,
    arena,
    attendance,
    plus_minus,
    clark = is_cc
  ) |>
  slice_sample(n = 10) |>
  janitor::clean_names(case = "title") |>
  gt() |>
  tab_header(
    title = "WNBA Game Data (1997-2025)",
    subtitle = "Sample of game-level records"
  ) |>
  fmt_number(
    columns = Attendance,
    decimals = 0
  ) |>
  tab_options(
    table.font.size = px(11),
    table.width = pct(100)
  )
```


@fig-boxplots shows the distribution of attendance across all seasons, with a logarithmic scale on the vertical axis. 
There is no attendance data from 2020, because all games that year were played in the WNBA bubble due to the COVID-19 pandemic, and some games in 2021 continued to be played with artificially low attendance.  

Several broad observations can be made from @fig-boxplots. 
First, after some initial enthusiasm during the first three years of the league, attendance largely plateaued or steadily but slightly declined from 2000 to 2017. 
The two years preceding the pandemic (2018-2019) had notably lower attendance. 
Since the pandemic, attendance has rebounded, exceeding its initial levels in 2024 and reaching new heights in 2025. 
These last two years coincide with the arrival of Caitlin Clark, the 2024 Rookie of the Year Award winner who is widely regarded as a generational talent. 
Disentangling the "Caitlin Clark effect" from the league's overall attendance trends is one of our goals. 

```{r}
#| label: fig-boxplots
#| fig-cap: "Distribution of WNBA game attendance by season. Note that the vertical axis has a logarithmic scale. Caitlin Clark's games in 2024-2025 are separated by color. "
ggplot(wnba_gl, aes(x = factor(season_id), y = attendance, color = factor(is_cc))) +
  geom_boxplot(outlier.alpha = 0.2, linewidth = 0.6, fill = "grey90") +
  geom_jitter(alpha = 0.2, width = 0.15, size = 1) +
  labs(
    title = "Distribution of WNBA Game Attendance by Season",
    subtitle = "1997-2025",
    x = "Season",
    y = "Attendance"
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Clark") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We used the game-level results data to model team strength as a continuous variable. 
Measuring team strength is necessary to test our hypothesis that attendance is in part driven by the desire of fans to see quality basketball. 
We considered season-to-date winning percentage and a rolling 44-game (the length of a typical WNBA season) winning percentage before settling on Elo rating [@Elo1978].
Originally developed for rating chess players, Elo rating is well-understood in sports analytics as a measure of team strength that updates after each game, and---unlike winning percentage and rolling winning percentage---takes into account not only the outcome but also the estimated team strength of both teams. 
We used a parameter value of $K = 32$ in our Elo ratings. 
@fig-elo shows the evolution of Elo rating by franchise for the 12 teams who were active during the 2024 season. 
The recent poor performance of the Los Angeles Sparks, the 2010 dominance of the Seattle Storm, and the recent rise of the New York Liberty and Indiana Fever are easily visible.

```{r}
#| label: fig-elo
#| fig-cap: "Evolution of team Elo ratings in the WNBA (1997-2025). Only franchises active in 2024 are shown. The recent poor performance of the Los Angeles Sparks, the 2010 dominance of the Seattle Storm, and the recent rise of the New York Liberty and Indiana Fever are easily visible. "
wnba_gl |>
  filter(is_active, team_abbreviation != "GSV") |>
  ggplot(aes(x = game_date, y = elo.A)) +
  geom_line() +
  facet_wrap(vars(name_current)) +
  scale_y_continuous("Elo rating") +
  scale_x_date("Season") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Methods

We included a 12-point home court advantage to our home win probability model. 

## Results


```{r}
#| label: tbl-model
#| tbl-cap: "Regression results for attendance model"
mod_attendance <- lm(
  log(attendance + 1) ~ home_win_prob + factor(wday(game_date)) + factor(month(game_date)) + name_current + factor(season_id) + is_cc,
  data = filter(wnba_gl, !season_id %in% c(2021), is_active)
)

models <- list(
  "OLS Model" = mod_attendance
)

modelsummary::modelsummary(
  models,
#  shape = model ~ term,
  output = "gt",
  statistic = "({std.error})",
  stars = TRUE,
  gof_omit = "IC|Log|Adj|Pseudo"
)
```


## Conclusion


## References {.unnumbered}

:::{#refs}

:::
